---
title: "metabolomics_analysis"
author: 
- "ddedesener"
- "DeniseSl22"
date: "25/04/2022"
output:
 md_document:
    variant: markdown_github
always_allow_html: true
---

## Introduction
In this workflow, we will apply statistical analysis on metabolomics data and link the metabolites of interest to pathway data from WikiPathways.

First we download the required data.
```{r data_download,warning=FALSE, message=FALSE}
# Set Working Directory, download data and store in /Data folder
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
work_DIR <- getwd()

# Obtain a file locally by its url
install.packages("downloader")
fileUrl <- "https://ibdmdb.org/tunnel/products/HMP2/Metabolites/1723/HMP2_metabolomics.csv.gz?accessType=DOWNLOAD"
require(downloader)
download(fileUrl, "data/metabolomics.csv.gz", mode = "wb")

#Note: if the URL download does not work, the zipped file is located on GitHub to continue the rest of this script.
if(!"R.utils" %in% installed.packages()){install.packages("R.utils")}
library(R.utils)
gunzip("data/metabolomics.csv.gz", remove=FALSE)

remove(fileUrl)
```

Second, we perform data extraction from the file, and process the data
```{r data_import,warning=FALSE, message=FALSE}
##DATA CLEANUP:

# Read in the csv file with data (note this might take some time, 81867 rows in original dataset)
mSet <- read.csv("data/metabolomics.csv", na.strings=c("", "NA"))

#Replace all empty values with NA
mSet[mSet==""] <- NA

#remove all lines with an empty value for the column called "HDMB"
mSet_Cleaned <- mSet[!is.na(mSet$"HMDB...Representative.ID."),]

#Remove all elements without an actual HMDB ID:
mSet_Cleaned_Ion <- mSet_Cleaned[!grepl("redundant ion", mSet_Cleaned$HMDB...Representative.ID.),]

if(!"dplyr" %in% installed.packages()){install.packages("dplyr")}
library(dplyr)

#IDs ending with a * (indicating ...) are placed in a separate column (to keep track of for future reference)
mSet_Cleaned_Star <- mSet_Cleaned_Ion%>%mutate(Star = case_when(
 grepl("[*]",mSet_Cleaned_Ion$HMDB...Representative.ID.) ~ 1
))

#Update IDs with a * to regular IDs.
mSet_Cleaned_HMDB <- data.frame(lapply(mSet_Cleaned_Star, function(x) {
                  gsub("[*]", "", x)
             }))

#Update IDs To new HMDB structure
mSet_Cleaned_HMDB_NEW <- data.frame(lapply(mSet_Cleaned_HMDB, function(x) {
                  gsub("HMDB", "HMDB00", x)
             }))


remove(mSet,mSet_Cleaned, mSet_Cleaned_Ion, mSet_Cleaned_Star, mSet_Cleaned_HMDB)

##DATA PROCESSING:

#Remove metabolites with < 50% data (located in columns 8-553.
removeLines <- rowSums(is.na(mSet_Cleaned_HMDB_NEW[,8-553]))
fifty_percent <- floor((553-8)/2)

mSet_MissingDataCounted <- cbind(mSet_Cleaned_HMDB_NEW, removeLines)
mSet_NoMissingData <- subset(mSet_MissingDataCounted, removeLines <= fifty_percent)

#Convert intensity data to numeric values                         
mSet_NoMissingData[ , c(8:553)] <- apply(mSet_NoMissingData[ , c(8:553)], 2, function(x) as.numeric(as.character(x)))

##TODO ask Duygu what this step does?
#set the rest features as column mean 

##normalization (see https://doi.org/10.1177%2F1469066720918446 and https://www.statology.org/transform-data-in-r/)
##Users can select different transformation styles here:
transformation <- "log_2" #log_2 transformation on intensity data selected for IBD dataset, options are: cube_root, square_root, log_2, log_10

if(transformation == "cube_root"){
    mSet_transformed <- cbind(mSet_NoMissingData[,c(1:7,554,555)], mSet_NoMissingData[,8:553]^(1/3))
}else if(transformation == "square_root"){
    mSet_transformed <- cbind(mSet_NoMissingData[,c(1:7,554,555)], mSet_NoMissingData[,8:553]^(1/2))
}else if(transformation == "log_2"){
    mSet_transformed <- cbind(mSet_NoMissingData[,c(1:7,554,555)], log2(mSet_NoMissingData[,8:553]))
}else if(transformation == "log_10"){
    mSet_transformed <- cbind(mSet_NoMissingData[,c(1:7,554,555)], log10(mSet_NoMissingData[,8:553]))
}else{print("Warning: name for transformation not recognized")}

## Visualize the data after the transformation (for one sample to get an idea of suitability of transformation:
#create histogram for original distribution for first column with data
hist(mSet_NoMissingData$HSM5FZBQ, col='steelblue', main='Original')
#create histogram for log-transformed distribution 
hist(mSet_transformed$HSM5FZBQ, col='coral2', main=transformation)

## Testing if the transformation creates a normally distributed dataset (alpha >= 0.05)
##Calculate all Shapiro values for raw and transformed data:
mSet_NoMissingData_Shapiro <- lapply(mSet_NoMissingData[,8:553], shapiro.test)
mSet_transformed_Shapiro <- lapply(mSet_transformed[,10:555], shapiro.test)

#Obtain the p-values for raw and transformed data
mSet_NoMissingData_Shapiro_pvalues <- do.call(rbind, mSet_NoMissingData_Shapiro)
mSet_transformed_Shapiro_pvalues <- do.call(rbind, mSet_transformed_Shapiro)

## Count how often the p-value is above 0.05, to obtain an estimate of achieved normality due to transformation
mSet_NoMissingData_Shapiro_pvalues_sum <- sum(mSet_NoMissingData_Shapiro_pvalues[,2] >= 0.05, na.rm=TRUE)
mSet_transformed_Shapiro_pvalues_sum <- sum(mSet_transformed_Shapiro_pvalues[,2] >= 0.05, na.rm=TRUE)

eighty_percent <- floor(((553-8)/10)*8)

#Print relevant information:
if(mSet_transformed_Shapiro_pvalues_sum[1] > eighty_percent ){paste0("Data after ", transformation ," transformation seems to follow a normal distribution for more then 80% of your data")} else{
  print("Advised to select a different data transformation procedure")}

remove(mSet_Cleaned_HMDB_NEW, mSet_MissingDataCounted, mSet_NoMissingData, mSet_NoMissingData_Shapiro, mSet_NoMissingData_Shapiro_pvalues, mSet_transformed_Shapiro, mSet_transformed_Shapiro_pvalues, eighty_percent, fifty_percent, mSet_NoMissingData_Shapiro_pvalues_sum, mSet_transformed_Shapiro_pvalues_sum, removeLines)
```

## Calculate logFC and p-value
```{r basic_statistics, warning=FALSE, message=FALSE}
##Download metadata, extract metabolomics sample IDs, location and disorders.
fileUrl <- "https://ibdmdb.org/tunnel/products/HMP2/Metadata/hmp2_metadata.csv?accessType=DOWNLOAD"
require(downloader)
download(fileUrl, "data/hmp2_metadata.csv", mode = "wb")

#Read metadata
htxMeta <- read.csv("data/hmp2_metadata.csv")
#Filter out by data type as metabolomics
htxMeta <- htxMeta  %>% filter(htxMeta$data_type == "metabolomics")

#Filter out data by disorders CD, UC and nonIBD samples 
htxMeta <-htxMeta  %>% filter(  (htxMeta$diagnosis == "CD" ) 
                                |  (htxMeta$diagnosis == "UC") 
                                |  (htxMeta$diagnosis == "nonIBD") 
                              )
#filter out unused columns
htxMeta <- htxMeta %>% select(External.ID,Participant.ID,diagnosis)
# order sample names based on columns external.ID
htxMeta <- htxMeta[order(htxMeta$External.ID),]

#Obtain Sample IDs which are both in mSet data and metadata
names.use <- names(mSet_transformed)[(names(mSet_transformed) %in% htxMeta$External.ID)]

#filter out metabolomics based on names.use and create metabolomicsFINAL (only retaining HMDB ID intensity data)
mSet_FINAL <- mSet_transformed[, names.use]

#metabolomicsFINAL data are ordered based on column names to match samples between metabolomicsFINAL and sampleLabels
mSet_FINAL <- mSet_FINAL[,order(names(mSet_FINAL))]
#check whether they are in same order
#colnames(mSet_FINAL) == htxMeta[,"External.ID"]

#add updated column names to the metadata file
colnames(htxMeta) <- c("sampleID", "participantID","disease")
#check whether sample names are in same order
checker <- all(colnames(mSet_FINAL) == rownames(htxMeta$sampleID)) 
paste("Matching between sample IDs and disorders went correctly:", checker)

#Connect disease info to metabolomics data
colname_Disorders <- htxMeta$disease
colnames(mSet_FINAL) <-c(colname_Disorders)

##Order columns based on disease
mSet_FINAL <- mSet_FINAL[ , order(names(mSet_FINAL))]

##ADD HMDB column at start
mSet_FINAL <- cbind(mSet_transformed$HMDB...Representative.ID., mSet_FINAL)
##Rename first column
colnames(mSet_FINAL)[1] <- "HMDB_ID"

#Find relevant columns per group.
library(stringr)
columns_CD <- sum(str_count(colname_Disorders, "CD"))
columns_nonIBD <- sum(str_count(colname_Disorders, "nonIBD"))
columns_UC <- sum(str_count(colname_Disorders, "UC"))

##Note: first column contains HMDB IDs.
end_CD_columns <- (columns_CD+1)
start_nonIBD_columns <- (columns_CD+2)
end_nonIBD_columns <- ((columns_CD+columns_nonIBD+1))
start_UC_columns <- ((columns_CD+columns_nonIBD+2))

##calculate logFC for 2 groups (CD vs control, UC vs control), ignoring missing values (NAs) when calculating the mean.  
disease_CD = apply(mSet_FINAL[,2:end_CD_columns], 1, mean, na.rm=TRUE)
control_IBD = apply(mSet_FINAL[,start_nonIBD_columns:end_nonIBD_columns], 1, mean, na.rm=TRUE)
disease_UC = apply(mSet_FINAL[,start_UC_columns:ncol(mSet_FINAL)], 1, mean, na.rm=TRUE)

#because the metabolomics data is already log2 transformed, we need to take the difference between the means (iso dividing the means over one another), since log2 Fold Change or log2 Ratio == log2(control / condition). Note: if the transformation step applied is cube_root or square_root, one needs to divide control over disease for this step!

if(transformation == "log_2" | transformation == "log10"){
    foldchange_CD <- control_IBD - disease_CD
    foldchange_UC <- control_IBD - disease_UC
}else{
  foldchange_CD <- (control_IBD / disease_CD)
  foldchange_UC <- (control_IBD / disease_UC)}

##ADD HMDB column at start, add fold change columns.
mSet_AnalysisReady <- cbind(mSet_FINAL$HMDB_ID, foldchange_CD, foldchange_UC)
##Rename first column
colnames(mSet_AnalysisReady)[1] <- "HMDB_ID"

##Calculate p-value for two groups based on t-test (comparing control to disease).
##general function to store p-values for multiple rows:
ttest_mSet <- function(df, grp1, grp2) {
  x = df[grp1]
  y = df[grp2]
  x = as.numeric(x)
  y = as.numeric(y)  
  results = t.test(x, y)
  results$p.value
}
##CD:
p_values_CD <- apply(mSet_FINAL, 1, ttest_mSet, grp1 = c(start_nonIBD_columns:end_nonIBD_columns), grp2 = c(2:end_CD_columns))
p_values_UC <- apply(mSet_FINAL, 1, ttest_mSet, grp1 = c(start_nonIBD_columns:end_nonIBD_columns), grp2 = c(start_UC_columns:ncol(mSet_FINAL)))

##Add p_values columns to analysis dataset:
mSet_AnalysisReady <- cbind(mSet_AnalysisReady, p_values_CD, p_values_UC)

#Convert logFC and p-values columns to numeric values            
mSet_AnalysisReady <- as.data.frame(mSet_AnalysisReady)
mSet_AnalysisReady[ , c(2:5)] <- apply(mSet_AnalysisReady[ , c(2:5)], 2, function(x) as.numeric(as.character(x)))

remove(fileUrl, mSet_transformed, names.use, transformation, columns_CD, columns_nonIBD, columns_UC, end_CD_columns, start_nonIBD_columns, end_nonIBD_columns, start_UC_columns, foldchange_CD, foldchange_UC, p_values_CD, p_values_UC, colname_Disorders, control_IBD, disease_CD, disease_UC, ttest_mSet, checker, htxMeta, mSet_FINAL)
```


## Volcano plot
--> Update script below, to create an Volcano plot without MetaboAnalystR

```{r volcano_plot,warning=FALSE, message=FALSE}
#See https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
##Potentially relevant packages needing to be updated before installing EnHancedVolcano (for Linux Debian):
install.packages("libfontconfig1-dev", "pkgconfig", "systemfonts", "Cairo")


#'getOption("repos")' replaces Bioconductor standard repositories, see '?repositories' for details

if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
  BiocManager::install('EnhancedVolcano', lib = "/home/deniseslenter/R/x86_64-pc-linux-gnu-library/4.1")

library(EnhancedVolcano)
  
EnhancedVolcano(mSet_AnalysisReady,
    lab = rownames(mSet_AnalysisReady),
    x = 'log2FoldChange',
    y = 'pvalue')

```


Export the data:
```{r data_export, warning=FALSE, message=FALSE}
# Saving output
if(dir.exists("output"))#if the output folder already exist
  unlink("output", recursive=TRUE)#first delete the existing one
dir.create("output")#create a new output folder
```

