---
title: "metabolomics_analysis"
author: 
- "DeniseSl22"
- "ddedesener"
date: "25/04/2022"
output:
 md_document:
    variant: markdown_github
always_allow_html: true
---

## Introduction
In this workflow, we link the metabolites of interest to pathway data from WikiPathways, based on their HMDB identifier.

```{r data_import,warning=FALSE, message=FALSE}
# Obtain Working Directory for step 8 to find processed data
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
getwd()
setwd('..')
work_DIR <- getwd()

#Obtain data from step 8
mSet_CD <- read.csv("8-significantly_changed_metabolites_analysis/output/mbxData_CD.csv", na.strings=c("", "NA"))
mSet_UC <- read.csv("8-significantly_changed_metabolites_analysis/output/mbxData_UC.csv", na.strings=c("", "NA"))

# Set Working Directory back to current folder
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
work_DIR <- getwd()

## Select a disorder to analyse (options; CD or UC)
disorder <- "CD"

if (disorder == "CD") {
  mSet = mSet_CD 
  print("Selected disorder is Crohn's disease")}else if(disorder == "UC"){ 
    mSet = mSet_UC
    print("Selected disorder is Ulcerative Colitis")}else{print("Disorder not Recognised")}
```
Find pathways based on relevant IDs column
```{r pathway_retrieval,warning=FALSE, message=FALSE}
if(!"SPARQL" %in% installed.packages()){
  install.packages("SPARQL")
}
library(SPARQL)
##Connect to Endpoint WikiPathways
endpointwp <- "https://sparql.wikipathways.org/sparql"
## 1. Query metadata:
queryMetadata <-
"SELECT DISTINCT ?dataset (str(?titleLit) as ?title) ?date ?license 
WHERE {
   ?dataset a void:Dataset ;
   dcterms:title ?titleLit ;
   dcterms:license ?license ;
   pav:createdOn ?date .
 }"
 #below code should be performed first to handle the ssl certificate error
options(RCurlOptions = list(cainfo = paste0( tempdir() , "/cacert.pem" ), ssl.verifypeer = FALSE))
resultsMetadata <- SPARQL(endpointwp,queryMetadata,curl_args=list(useragent=R.version.string))
showresultsMetadata <- resultsMetadata$results
remove(queryMetadata, resultsMetadata)


## Create a list of HMDB IDs according to filtering criteria from step 8.
list_Relevant_HMDB_IDs <- list(mSet$relevant_ids)
vector_HMDB <- unlist(list_Relevant_HMDB_IDs) #convert list to array, for traversing the data to a SPARQL query later on
vector_HMDB <- vector_HMDB[!is.na(vector_HMDB)]
##Add the HMDb prefix IRI in front of all IDs.
query_HMDBs <- paste("ch:", vector_HMDB, sep="")
##Merge the individual entries in the vector into one string, separated by a space
string_HMDB <- paste(c(query_HMDBs), collapse=' ' )

##TODO: add column with nr. of Metabolites in PW (to calculate PW impact)

#For now, filter out Reactome PWs due to visualization issues in Cytoscape.
item1 = "PREFIX ch: <https://identifiers.org/hmdb/>
PREFIX cur: <http://vocabularies.wikipathways.org/wp#Curation:>
select distinct ?pathwayRes (str(?wpid) as ?pathway) (str(?title) as ?pathwayTitle) (count(distinct ?hmdbMetabolite) AS ?HMDBsInPWs) (count(distinct ?metaboliteDatanode) AS ?TotalMetabolitesinPW) (GROUP_CONCAT(DISTINCT fn:substring(?hgnc,37);separator=' ') AS ?Proteins) (count(distinct ?hgnc) AS ?ProteinsInPWs) (GROUP_CONCAT(DISTINCT fn:substring(?hmdbMetabolite,30);separator=' ') AS ?includedHMDBs)
where {
VALUES ?hmdbMetabolite {"
item2 = "}
 
 ?metaboliteDatanode	a wp:Metabolite ;
                        dcterms:isPartOf ?pathwayRes .
 
 ?datanode	a wp:Metabolite ;          
           	wp:bdbHmdb  ?hmdbMetabolite ;
    		dcterms:isPartOf ?pathwayRes .
 ?pathwayRes a wp:Pathway ;
             wp:organismName 'Homo sapiens' ; 
    		dcterms:identifier ?wpid ;
    		dc:title ?title .
    		
 ?datanode2 wp:bdbHgncSymbol ?hgnc ;
    		dcterms:isPartOf ?pathwayRes .
    		
  #?pathwayRes wp:ontologyTag cur:Reactome_Approved . 
  ?pathwayRes wp:ontologyTag cur:AnalysisCollection .   		
}
ORDER BY DESC(?HMDBsInPWs)"
query_CombinePWs <- paste(item1,string_HMDB,item2)
remove(item1, item2)

results_CombinePWs <- SPARQL(endpointwp,query_CombinePWs,curl_args=list(useragent=R.version.string))
showresults_CombinePWs <- results_CombinePWs$results
remove(query_CombinePWs,results_CombinePWs)
#Print table with first 5 relevant pathways (if less than 5 are found, print only those)
if(nrow(showresults_CombinePWs) < 5){
print(showresults_CombinePWs[1:nrow(showresults_CombinePWs),c(2:5)])
}else{print(showresults_CombinePWs[1:5,c(2:5)])}

remove(cleaned_string_HMDB, list_Relevant_HMDB_IDs, query_HMDBs)
```
Calculate the ORA score for each pathway, using the Fishers exact test.
```{r pathway_analysis, warning=FALSE, message=FALSE}
##Based on: https://www.pathwaycommons.org/guide/primers/statistics/fishers_exact_test/
##TODO: finish this section!

#Create a dataframe to store the required numbers in.
Contingency_table <- data.frame(matrix(ncol=5,nrow=0, dimnames=list(NULL, c("WP.ID", "x", "m", "n", "k"))))
counter = 1
for (i in 1:nrow(showresults_CombinePWs)) {
   Contingency_table[counter,1] <- (showresults_CombinePWs[i,2]) #WP.ID
   Contingency_table[counter,2] <- (showresults_CombinePWs[i,4]) ##x <- (number4) #Total differentially changed metabolites, also in a PW. (HMDBsInPWs)
   Contingency_table[counter,3] <- (showresults_CombinePWs[i,5]) ##m <- (number) #Total Metabolites in PW (TotalMetabolitesinPW)
   Contingency_table[counter,4] <- (length(unique(mSet[,1])) - showresults_CombinePWs[i,4]) ##n <- (number2) #Total Metabolites measured not in PW (DISTINCT all_HMDB - HMDBsInPWs)
   Contingency_table[counter,5] <- length(unique(vector_HMDB)) ##k <- (number3) #Total differentially changed metabolites. (DISTINCT vector_HMDB)

   counter <- counter + 1
}

# Calculate hypergeometric density p-value for all pathways.
i <- 1:nrow(Contingency_table)
probabilities <- dhyper(Contingency_table[i,2], Contingency_table[i,3], Contingency_table[i,4], Contingency_table[i,5], log = FALSE)

pathwayAnalysis_results <- cbind(showresults_CombinePWs[, c(2:4)], probabilities, showresults_CombinePWs[, c(6,7)])
colnames(pathwayAnalysis_results)[5] <- "HGNCs"
colnames(pathwayAnalysis_results)[6] <- "ProteinsInPWs"

##Sort PW results based on 1. highest amount of #HMDBs in PW, 2. lowest p-values,  3. highest amouny of proteins in PW (which might be relevant for transcriptomics analysis later)
pathwayAnalysis_results_sorted <- pathwayAnalysis_results[  with(pathwayAnalysis_results, order(-HMDBsInPWs, probabilities, -ProteinsInPWs)),]

print(pathwayAnalysis_results_sorted[1:5,])
```
Export the pathway data:
```{r data_export, warning=FALSE, message=FALSE}
##Save the data file
nameDataFile <- paste0("output/mbxPWdata_", disorder ,".csv")
write.table(pathwayAnalysis_results_sorted, nameDataFile, sep =",", row.names = FALSE)
```

Print significantly changed metabolites which were not in a pathway, by ID and name:
```{r missing_data,warning=FALSE, message=FALSE }
##Find Missing Biomarkers (not part of any Human pathway model)
item1 = "PREFIX ch: <https://identifiers.org/hmdb/>
SELECT DISTINCT ?HMDBMetabolite WHERE {
  VALUES ?HMDBMetabolite {"
item2 = "}
  ?pathwayRes  a wp:Pathway ;
             	wp:organismName 'Homo sapiens' .
  
  ?metabolite 	a wp:Metabolite ;
                dcterms:identifier ?id ;
                dcterms:isPartOf ?pathwayRes .
  ?metabolite wp:bdbHmdb ?HMDBMetabolite.
}"
queryMissingBiomarkers <- paste(item1,string_HMDB,item2)
remove(item1,item2)
resultsMissingBiomarkers <- SPARQL(endpointwp,queryMissingBiomarkers,curl_args=list(useragent=R.version.string))
listMissingBiomarkers <- c(resultsMissingBiomarkers$results) #safe results as list for comparison.
remove(queryMissingBiomarkers,resultsMissingBiomarkers)
HMDBs_inPWs <- gsub("[<https://identifiers.org/hmdb/>]", "", listMissingBiomarkers) #HMDB IDs IRI cleanup
intersectingHMDB <- setdiff(vector_HMDB, HMDBs_inPWs)

string_intersectingHMDB <- paste(c(intersectingHMDB), collapse=', ' )

#Find names for missing Biomarkers based on HMDB ID (to help with data understanding and curation)
missingNames <- list()
for (j in 1:length(intersectingHMDB)){
  for (i in 1:nrow(mSet)){
    if(!is.na(mSet[i,5]) & mSet[i,5] == intersectingHMDB[j]){
       missingNames[j] <- mSet[i,6]
      }
    else{next}
  }
}
remove(i,j)
#Save list on one string for reporting purposes
string_missingNames <- do.call(paste, c(as.list(missingNames), sep = ", "))
#Print relevant information:
if(length(intersectingHMDB) == 0 ){print("All relevant biomarkers are in a pathway!")} else{
  print(paste0("For the disorder ", disorder, ", ", length(intersectingHMDB), " biomarkers are not in a pathway; with the following HMDB IDs: " , string_intersectingHMDB, "; with the following Database names: ", string_missingNames))}
```
### Last, we create a Jupyter notebook and markdown file from this script
```{r writing_to_notebooks,warning=FALSE, message=FALSE }
#Jupyter Notebook file
if(!"devtools" %in% installed.packages()) BiocManager::install("devtools")
devtools::install_github("mkearney/rmd2jupyter", force=TRUE)
library(devtools)
library(rmd2jupyter)
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
rmd2jupyter("metabolomics_pathway_analysis.Rmd")

#markdown_file <- "metabolomics_pathway_analysis.md"
#if (file.exists(markdown_file)) {
#   unlink(markdown_file, recursive=TRUE)#first delete the existing one
# }
#If this next line trows an error, build the md file with knittr manual selection (file, Knit document, or ctrl+shift+k -keyboard shortcut).
#rmarkdown::render("metabolomics_pathway_analysis.Rmd", "md_document")

##Clean up data
remove(counter, i, probabilities, string_HMDB, vector_HMDB, Contingency_table, pathwayAnalysis_results, showresults_CombinePWs, showresultsMetadata)
```
