---
title: "metabolomics_analysis"
author: "ddedesener"
date: "03/01/2022"
output:
 md_document:
    variant: markdown_github
always_allow_html: true
---

## Introduction
In this workflow, we will apply statis one factoral analysis on metabolomics data
using MetaboAnalyst software R packages.
MetaboAnalystR 3 contains the R functions and libraries underlying the popular MetaboAnalyst web server, including metabolomic data analysis, visualization, and functional interpretation. The package is synchronized with the MetaboAnalyst web server. After installing and loading the package, users will be able to reproduce the same results from their local computers using the corresponding R command history downloaded from MetaboAnalyst web site, thereby achieving maximum flexibility and reproducibility.
source:https://www.metaboanalyst.ca/MetaboAnalyst/docs/RTutorial.xhtml

## Setup
Installing and loading required libraries
```{r setup_environment, warning=FALSE, message=FALSE}
# check if libraries are already installed > otherwise install it
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if(!"rstudioapi" %in% installed.packages()) BiocManager::install("rstudioapi")
if(!"MetaboAnalystR" %in% installed.packages()) BiocManager::install("MetaboAnalystR")

#load installed librariers
library(BiocManager)
library(rstudioapi)
library(MetaboAnalystR)
```

# Data import and preprocessing
```{r data_import,warning=FALSE, message=FALSE}
# set your working environment to the location where your current source file is saved into.
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
work_DIR <- getwd()

#initialize data object
mSet<-InitDataObjects("pktable", "stat", FALSE)
#read data
mSet<-Read.TextData(mSet, "data/mbxDataCD_nonIBD.csv", "colu", "disc");
mSet<-SanityCheckData(mSet)

#missing value removal by filtering out features has more than 50% missing
mSet<-RemoveMissingPercent(mSet, percent=0.5)
#set the rest features as column mean
mSet<-ImputeMissingVar(mSet, method="mean")
mSet<-SanityCheckData(mSet)

#no filter applied
mSet<-FilterVariable(mSet, "none", "F", 25)

# Saving output
if(dir.exists("output"))#if the output folder already exist
  unlink("output", recursive=TRUE)#first delete the existing one
dir.create("output")#create a new output folder
#setwd(paste0(work_DIR,"/output"))

#normalization
mSet<-PreparePrenormData(mSet)
mSet<-Normalization(mSet, "NULL", "CrNorm", "NULL", ratio=FALSE, ratioNum=20)
mSet<-PlotNormSummary(mSet, "norm_0_", "png", 72, width=NA)
mSet<-PlotSampleNormSummary(mSet, "snorm_0_", "png", 72, width=NA)
```

## Volcano plot
```{r volcano plot,warning=FALSE, message=FALSE}
mSet<-Volcano.Anal(mSet, FALSE, 1.0, 0, F, 1.0, TRUE, "raw")
mSet<-PlotVolcano(mSet, "volcano_1_",1, 0, "png", 72, width=NA)
```

